---
title: "HW 2"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(dplyr)
```


```{r }
#2
classes <- c('character', 'character', 'factor', 'character', 'double', 'integer', 'integer', 'integer', 'integer', 'integer', 'integer', 'integer', 'integer', 'integer', 'integer', 'integer', 'integer', 'integer', 'integer', 'integer', 'integer', 'integer', 'integer', 'integer')
#base method
dat <- read.csv('nba2017-player-statistics.csv', colClasses = classes)
str(dat)
#readr method
datt <- read_csv('nba2017-player-statistics.csv', col_types =  list(
   Player = col_character(),
   Team = col_character(),
   Position = col_factor(c('C', 'PF', 'PG', 'SF', 'SG')),
   Experience = col_character(),
   Salary = col_double(),
   Rank = col_integer(),
   Age = col_integer(),
   GP = col_integer(),
   GS = col_integer(),
   MIN = col_integer(),
   FGM = col_integer(),
   FGA = col_integer(),
   Points3 = col_integer(),
   Points3_atts = col_integer(),
   Points2 = col_integer(),
   Points2_atts = col_integer(),
   FTM = col_integer(),
   FTA = col_integer(),
   OREB = col_integer(),
   DREB = col_integer(),
   AST = col_integer(),
   STL = col_integer(),
   BLK = col_integer(),
   TO = col_integer()
))
str(datt)

#3
datt$Experience[datt$Experience == "R"] <- 0 # Replace "R" with '0'
datt$Experience
as.numeric(as.character(datt$Experience)) #change Experience column from char to numbers


#4
Missed_FG <- datt$FGA - datt$FGM
Missed_FT <- datt$FTA - datt$FTM
PTS <- (datt$Points3 * 3) + (datt$Points2 * 2) + (datt$FTM * 1)
MPG <- datt$MIN / datt$GP
REB <- datt$OREB + datt$DREB
datt <- cbind(datt, Missed_FG)
datt <- cbind(datt, Missed_FT)
datt <- cbind(datt, PTS)
datt <- cbind(datt, MPG)
datt <- cbind(datt, REB)
EFF <- (datt$PTS + datt$REB + datt$AST + datt$STL + datt$BLK - datt$Missed_FG - datt$Missed_FT - datt$TO) / datt$GP
datt <- cbind(datt, EFF)

summary(datt$EFF)

#histogram
#hist(c(EFF, table(EFF)))
hist(EFF, freq = table(EFF))
#display player, team, salary, eff for top 10 players by eff in decreasing order
dat2 <- select(datt, Player, Team, Salary, EFF)
head(arrange(dat2, desc(datt$EFF)), 10)

#filter players with negative EFF
filter(dat2, EFF < 0)
#cor of EFF with other variables
xpts <- cor(EFF, datt$PTS)
xpts
xreb <- cor(EFF, datt$REB)
xreb
xast <- cor(EFF, datt$AST)
xast
xstl <- cor(EFF, datt$STL)
xstl
xblk <- cor(EFF, datt$BLK)
xblk
xmg <- cor(EFF, -datt$Missed_FG)
xmg
xmt <- cor(EFF, -datt$Missed_FT)
xmt
xto <- cor(EFF, -datt$TO)
xto
xgp <- cor(EFF, datt$GP)
xgp

bpdata <- c(xpts, xreb, xast, xstl, xblk, xmg, xmt, xto, xgp)
newD <- sort(bpdata, decreasing = TRUE)
newD
#barplot
barplot(newD, col = c("blue", "blue", "blue", "blue", "blue", "blue", "red", "red", "red"), main = 'Correlation Between Player Stats and EFF',names.arg = c('PTS', 'REB', 'AST', 'STL', 'BLK', 'MG', 'MT', 'TO', 'GP'), border =TRUE, ylim = c(-1,1))

#5
#plot
plot(datt$EFF, datt$Salary/1000000,col = "#55555555" ,cex = .8,pch = 19, xlab = "EFF (efficiency)", ylab = "Salary (in millions)", main = "Scatterplot with lowess smooth")
lines(lowess(datt$EFF, datt$Salary/1000000), lwd = 3, col = "#F15C3C")

#linear correlation coefficient
n <- length(datt$Salary)
n
salary_mean = mean(datt$Salary)
salary_mean

eff_mean = mean(datt$EFF)
eff_mean

salary_var = var(datt$Salary)
salary_var

eff_var = var(datt$EFF)
eff_var

covES <- cov(datt$Salary, datt$EFF)
covES

correlation <- covES / (sqrt(salary_var) * sqrt(eff_var))
correlation

# What can you say about relationship between the two? It appears that there is a positive correlation between salary and eff score, the higher the eff score, it is probably the higher the salary since it would mean, in theory, the player is performing at a good level.

#Rookie Calculation

newMPG <- filter(datt, datt$MPG >= 20)
players2 <- data.frame(newMPG)
players2

#scatterplot
plot(players2$EFF, players2$Salary/1000000,col = "#55555555" ,cex = .8, xlab = "EFF (efficiency)", ylab = "Salary (in millions)", main = "Scatterplot with lowess smooth")
lines(lowess(players2$EFF, players2$Salary/1000000), lwd = 3, col = "#F15C3C")

#linear correlation coefficient
n <- length(players2$Salary)
n
salary_mean = mean(players2$Salary)
salary_mean

eff_mean = mean(players2$EFF)
eff_mean

salary_var = var(players2$Salary)
salary_var

eff_var = var(players2$EFF)
eff_var

covES <- cov(players2$Salary, players2$EFF)
covES

correlation <- covES / (sqrt(salary_var) * sqrt(eff_var))
correlation

#What can you say about the relationship now? The correlation is less positie than before, it does not seem that a good EFF score implies a higher salary. There are a lot of factors with salary caps that contribute to this not being as positive as before. More experience does not necessarily mean earn more.

```

#### 6: Comments and Reflections
- What things were hard, even though you saw them in class/lab?
  1. Making the plots look nice, figuring out read_csv()
- What things were easy even though we haven't done it in class/lab?
  1. Some of the data frame manipulation
- Did you need help to complete the assignment>
  1. No.
- How much time did it take to complete this hw?
  1. A couple of hours, (3-4)
- What was the most time consuming part?
  1. The setting up of the read_csv()
- Was there anything you did not understand?
  1. No, it is all straight forward
- Was there anything frustrating in particular?
  1. A lot of the work is repetitive and tedious
- Was there anything exciting? Something you feel proud of?
  1. I'm learning R!
  




