---
title: "HW03"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source('/Users/Roberto/Desktop/133/stat133-hws-fall17/hw03/code/make-teams-table.R')
```

## R Markdown

```{r }

gg_salary <- ggplot(data = teams, aes(x =reorder(team, salary) ,y =salary, group = 1)) + labs(x= "Teams", y = "Salary") +
  geom_bar(stat='identity') + coord_flip() 
gg_salary <- gg_salary + geom_hline(yintercept = mean(teams$salary),color = "red")
gg_salary


gg_points <- ggplot(data = teams, aes(x =reorder(team, points) ,y =points, group = 1)) + labs(x= "Teams", y = "Points") +
  geom_bar(stat='identity') + coord_flip() 
gg_points <- gg_points + geom_hline(yintercept = mean(teams$points),color = "red")
gg_points

gg_eff <- ggplot(data = teams, aes(x =reorder(team, efficiency) ,y =efficiency, group = 1)) + labs(x= "Teams", y = "Points") +
  geom_bar(stat='identity') + coord_flip() 
gg_eff <- gg_eff + geom_hline(yintercept = mean(teams$efficiency),color = "red")
gg_eff


pcateam <- select(teams, points3, points2, free_throws, off_rebounds, def_rebounds, assists, steals, blocks, turnovers, fouls)

pca <- prcomp(pcateam, scale.=TRUE)
pca


eigenvalue = pca$sdev^2
prop = round(pca$sdev^2 / sum(pca$sdev^2), 4)
cprop = cumsum(prop)
eigs <- data.frame(
  eigenvalue,
  prop,
  cprop
)
eigs

pc1 <- pca$x[,1] 
pc2 <- pca$x[,2] 
plot(pc1, pc2, main = "Scatter of PC1 and PC2", cex = 0)
text(pc1,pc2,labels=teams$team)


s1 <- 100 * ((pc1 - min(pc1))/ max(pc1) - min(pc1))
gg_pc1 <- ggplot(data = teams, aes(x =reorder(teams$team, s1) ,y =s1, group = 1)) + labs(x= "Team", y = "First PC (scaled from 0 to 100") +
  geom_bar(stat='identity') + coord_flip() 
gg_pc1

```


Comments and Reflections:

Whas this your firt time working on a project with such file structure?
Yes, it was interested, but seems like too much work
Was this your first time using rel paths?
NO
Was this your first time using an R script?
NO
What things were hard?
Understanding PCA
What was easy ish?
The R script part
Did anyone help you?
No
How much time did it take to complete this hw?
About 3 hours
What was the most time consuming?
PCA part
Anything interesting?
The plots




